---
title: "Case Study 8: Building the past"
author: "Devin W. H."
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:  
    keep_md: true
    toc: true
    toc_float: true
    code_folding: hide
    fig_height: 6
    fig_width: 12
    fig_align: 'center'
---

```{r, echo=FALSE}

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

```

```{r load_libraries, include=FALSE}

devtools::install_github("hathawayj/buildings")
library(tidyverse)
library(USAboundaries)
library(buildings)
library(sf)
library(scales)

```

```{r load_data}

states <- us_states() %>%
  mutate(statefp = as.numeric(statefp))

counties <- us_counties(state = 'Arizona') %>%
  select(-state_name) %>%
  mutate(countyfp = as.numeric(countyfp))

# For states
single_fam_states <- permits %>%
  filter(variable == 'Single Family') %>%
  rename(statefp = state, countyfp = county)

# For counties in Alaska
single_fam_counties <- permits %>%
  filter(state == 4, variable == 'Single Family') %>%
  rename(statefp = state, countyfp = county)

```

## Background

The purpose of this analysis is to look back on single family building permits across the U.S over the years with a special case scenario glance at Arizona's counties. This is done with the intent on comparing the overall trends to the housing crisis event of 2008.

## Data Wrangling

```{r tidy_data}

state_permit <- left_join(single_fam_states, states, by = 'statefp') %>%
  filter(!StateAbbr %in% c('AK', 'HI')) %>%
  group_by(statefp, year) %>%
  summarise(geometry = geometry, value = value, year = year) %>%
  mutate(total = sum(value)) %>%
  unique()

counties_permit <- left_join(single_fam_counties, counties, by = 'countyfp')

total_permit <- single_fam_states %>%
  group_by(year) %>%
  summarise(total = sum(value))

```

## Data Visualization

States with the most permits are California, Texas, and Florida. Between 2004 and 2006 there is a noticeable change in coloration representative of number of housing permits for single families. 

```{r plot_data, fig.width = 20, fig.height = 20}

ggplot(state_permit) +
  aes(geometry = geometry, fill = total) +
  facet_wrap(~ year) +
  geom_sf() +
  scale_fill_gradient(
    low = "white",
    high = "firebrick",
    space = "Lab",
    na.value = "grey50",
    guide = "colourbar",
    aesthetics = "fill",
    name = "Number of Permits") +
  labs(title = 'Single Family Housing Permits by Continental States') +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5),
        legend.position = 'bottom')

```

Foor Arizona in particular, Maricopa county (the one I was born in) has the highest amount of building permits for single families compared to any other within the state. Since 1980 there was a steady increase each year until 2005 where we see it immediately start to fade out.

```{r fig.width = 20, fig.height = 20}

ggplot(counties_permit) +
  facet_wrap(~ year) +
  aes(geometry = geometry, fill = value) +
  geom_sf() +
  scale_fill_gradient(
    low = "white",
    high = "firebrick",
    space = "Lab",
    na.value = "grey50",
    guide = "colourbar",
    aesthetics = "fill",
    name = "Number of Permits") +
  labs(title = 'Single Family Housing Permits in Arizona Counties') +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5),
        legend.position = 'bottom')

```

The line graph details the total single family building permits in the country over the years. As plainly seen, there is a major drop off at the 2005 mark with the lowest point being around 2008, showing that the housing crisis climaxed on this year before an ever so slight increase into 2010. This graph showcases the significance of the crisis more than the maps.

```{r}

ggplot(total_permit) +
  aes(x = year, y = total) +
  geom_line(color = 'firebrick') +
  theme_minimal() +
  scale_y_continuous(labels = comma) +
  scale_x_continuous(breaks = seq(1980, 2010, 5)) +
  labs(x = 'Year', y = 'Total Permits', title = 'Total Permits In the U.S') +
  theme(plot.title = element_text(hjust = 0.5))

```

## Conclusions

When showcasing the national crisis of the housing market, using the line graph that obtains total permits for each year better showcases the massive drop point from 2005 onward. 
