---
title: "Case Study 6: It’s about time"
author: "Devin W. H."
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:  
    keep_md: true
    toc: true
    toc_float: true
    code_folding: hide
    fig_height: 6
    fig_width: 12
    fig_align: 'center'
---

```{r, echo=FALSE}

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

```

```{r load_libraries, include=FALSE}

library(tidyverse)
library(lubridate)
library(ggthemes)

```

```{r load_data}

data <- rio::import('https://byuistats.github.io/M335/data/sales.csv')

```

## Background

Record of transactions were obtained from several businesses in various industries. The purpose is to identify which business has performed the best within a three month period for a loan program. No forecasting will be conducted, but only an analysis of current time series on the revenue of several businesses from May to July.

## Data Wrangling

```{r tidy_data}

data <- data %>%
  mutate(Time = with_tz(Time, tzone = 'America/Denver')) %>%
  filter(!Name == 'Missing')

```

```{r}

# Provide visualizations that show gross revenue over time for each company (Choose if you want to aggregate at the daily, the weekly, or the monthly level).

sales1 <- data %>%
  mutate(Month = month(Time, label = TRUE), Year = year(Time)) %>%
  group_by(Name, Month) %>%
  summarise(Gross = sum(Amount)) %>%
  filter(!Month == 'Apr') # Only two instances of April transactions

```

```{r}

# Provide a visualization that gives insight into hours of operation for each company.

sales2 <- data %>%
  mutate(Hour = hour(Time)) %>%
  count(Name, Hour)

```

```{r}

# We don’t have employee numbers, but customer traffic (number of transactions) may be helpful. Provide a visualization on customer traffic for each company.

sales3 <- data %>%
  mutate(rnd_day = floor_date(Time, unit = 'day')) %>%
  count(Name, rnd_day)

```

## Data Visualization

The following graph shows *gross revenue* for each company monthly. June saw the biggest increase in profits from the previous months with July being slightly more profitable than June for LeBelle.

```{r plot_data}

# Provide visualizations that show gross revenue over time for each company (Choose if you want to aggregate at the daily, the weekly, or the monthly level).

ggplot(sales1) +
  aes(x = Month, y = Gross, color = Name) +
  facet_wrap(~Name) +
  geom_line(aes(group = Name), size = 1.5) +
  theme_hc() +
  theme(legend.position = 'None')

```

The graph below finds number of transactions for each store during each hour of the day. Most businesses find the most profitable times to be between 10 and 3 which is expected to be the busiest time frame regardless of industry. It isn't clear what hours the stores open or close since this only gathers transactions.

```{r}

# Provide a visualization that gives insight into hours of operation for each company.

ggplot(sales2) +
  aes(x = Hour, y = n, fill = Name) +
  facet_wrap(~ Name) +
  geom_col() +
  coord_cartesian(xlim = c(7, 24)) +
  theme_gdocs() +
  theme(legend.position = 'None')

```

Down below is daily transactions for customer traffic at each business. Frozone, LeBelle, SplashAndDash, and Tacontento stay mostly similar throughout the months while HotDiggity and Shortstop fluctuate more frequently.

```{r}

# We don’t have employee numbers, but customer traffic (number of transactions) may be helpful. Provide a visualization on customer traffic for each company.

ggplot(sales3) +
  aes(x = rnd_day, y = n, color = Name) +
  facet_wrap(~ Name, nrow = 1) +
  geom_point() +
  geom_smooth(color = 'black') +
  theme_calc() +
  theme(legend.position = 'None')

```

## Conclusions

It is apparent HotDiggity has the best sales of any of the analyzed businesses. However, it should be noted that their business seems to only spike during the summer which may be a risk when it comes to paying back on a loan. LeBelle and SplashandDash have more of a consistent source of income. Cross reference the bottommost graph with the topmost and it can be inferenced that SplashandDash has a good amount of income at a more consistent rate. Were I to give out a loan to expand a business, I would recommend SplashandDash as a good candidate along with LeBelle.
